---
title: "Applying Machine Learning Methods to Predict Stroke"
author: "Malvika Venkataraman (msv2137), Anand Rajan (ar4173), Benjamin Goebel (bpg2118)"
date: "May 8, 2022"
output: pdf_document
---

```{r}
library(tidyverse)
library(caret)
library(recipes)
library(kernlab)
library(ISLR)
library(mlbench)
library(ranger)
library(gbm)
library(pdp)
library(pROC)
```

# Introduction
## Data/Motivation
For our project our team chose to use the stroke prediction dataset from [kaggle](https://www.kaggle.com/datasets/fedesoriano/stroke-prediction-dataset?datasetId=1120859&language=R). [Stroke](https://www.stroke.org/en/about-stroke#:~:text=Stroke%20is%20a%20disease%20that,or%20bursts%20(or%20ruptures)) is a disease that affects the arteries leading to and within the brain. It is the No. 5 cause of death and a leading cause of disability in the United States. Our motivation was to use this dataset to help identify key indicators that lead to strokes, since many are preventable. The questions we are trying to answer using this dataset are: 1) What are the key indicators or risk factors that lead to stroke occurrence? 2) Which classification model performs best in predicting these key indicators?

## Data Cleaning
To properly analyze the data and build models to make predictions, the first step is to clean the data. The vast majority of the data was already fairly clean and ready for analysis, but we began by reading in the data, cleaning variable names, making categorical data factors, making continuous data of type numeric, and dropping missing values. We also decided to exclude the variable "id" since it will not be relevant to our analysis. 
```{r, warning = FALSE, include=FALSE}
# data cleaning
stroke_data <- read_csv("./data/healthcare-dataset-stroke-data.csv")
stroke_data <-  janitor::clean_names(stroke_data)

# Convert BMI to numeric
stroke_data$bmi <- as.numeric(stroke_data$bmi)

# change categorical variables to factors
stroke_data$gender <- ifelse(stroke_data$gender == "Other", "Female", stroke_data$gender)
stroke_data$gender = as.factor(stroke_data$gender)
stroke_data$hypertension = as.factor(ifelse(stroke_data$hypertension == 1, "Yes", "No"))
stroke_data$heart_disease = as.factor(ifelse(stroke_data$heart_disease == 1, "Yes", "No"))
stroke_data$ever_married = as.factor(stroke_data$ever_married)
stroke_data$work_type = as.factor(stroke_data$work_type)
stroke_data$residence_type = as.factor(stroke_data$residence_type)
stroke_data$smoking_status = as.factor(stroke_data$smoking_status)
stroke_data$stroke = as.factor(ifelse(stroke_data$stroke == 1, "Yes", "No"))
# remove id variable
stroke_data = stroke_data[,2:12]
# change from tibble to dataframe
stroke_data = as.data.frame(stroke_data)
```

The resulting dataset contains `r nrow(stroke_data)` patient records and `r ncol(stroke_data)` columns. The dependent variable is the binary variable "stroke" with response values "Yes" and "No". Other variables include: gender, age, hypertension, heart disease, ever been married, work type, residence type, average glucose level, BMI and smoking status.

# EDA
To create a graphical summary of the numeric data, we used featurePlot. Note that to produce the density plots for this dataset we need to exclude the categorical variables, so they were removed.
```{r, echo=FALSE}
theme1 <- transparentTheme(trans = .4)
trellis.par.set(theme1)

featurePlot(x = stroke_data[, c(2,8,9)],
            y = stroke_data$stroke, 
            scales = list(x = list(relation = "free"), 
                          y = list(relation = "free")), 
            plot = "density", 
            pch = "|",
            auto.key = list(columns = 2))
```

From the plots, we can see that the distribution of BMI and average glucose level for patients who experienced a stroke, versus patients who didn't are fairly similar. However, the distribution of age across the two groups is fairly different. Patients who experienced a stroke tended to be of a higher age.

To visualize the categorical variables, we decided to create plots of the relative proportion of variable levels faceted by stroke occurrence. These plots help visualize any noticeable differences in the distribution of the categorical variables by stroke type.

```{r, echo=FALSE}
gender_plot <- ggplot(stroke_data, aes(x = gender)) +  
  geom_bar(aes(y = ..prop.., group = 1), fill = 'blue') + labs(x = 'gender') + 
  facet_wrap(~ stroke) 

hypertension_plot <- ggplot(stroke_data, aes(x = hypertension)) +
geom_bar(aes(y = ..prop.., group = 1), fill = 'blue') +  
  labs(x = 'hypertension') + facet_wrap(~ stroke)

heart_disease_plot <- ggplot(stroke_data, aes(x = heart_disease)) +
geom_bar(aes(y = ..prop.., group = 1), fill = 'blue') +  
  labs(x = 'heart_disease') + facet_wrap(~ stroke)

ever_married_plot <- ggplot(stroke_data, aes(x = ever_married)) +
geom_bar(aes(y = ..prop.., group = 1), fill = 'blue') +  
  labs(x = 'ever_married') + facet_wrap(~ stroke)

work_type_plot <- ggplot(stroke_data, aes(x = work_type)) +
geom_bar(aes(y = ..prop.., group = 1), fill = 'blue') +  
  labs(x = 'work_type') + theme(axis.text.x = element_text(angle = 90)) + facet_wrap(~ stroke)

residence_type_plot <- ggplot(stroke_data, aes(x = residence_type)) +
geom_bar(aes(y = ..prop.., group = 1), fill = 'blue') +  
  labs(x = 'residence_type') + theme(axis.text.x = element_text(angle = 90)) + facet_wrap(~ stroke)

smoking_status_plot <- ggplot(stroke_data, aes(x = smoking_status)) +
geom_bar(aes(y = ..prop.., group = 1), fill = 'blue') +  
  labs(x = 'smoking_status') + theme(axis.text.x = element_text(angle = 90)) + facet_wrap(~ stroke)

residence_type_plot + hypertension_plot + heart_disease_plot + ever_married_plot + plot_layout(ncol = 2) + 
  plot_layout(widths = c(.1, .1), heights = c(.1, .1)) + plot_annotation(title = "Relative Proportion of Variables By Stroke Occurence")

work_type_plot + smoking_status_plot + gender_plot + plot_layout(ncol = 2)

```

From the visualizations, we can see that among those who experienced strokes: a noticeably higher proportion of individuals had hypertension, heart disease, have ever been married, have formerly smoked or smoke currently. These variables could be potential indicators of stroke occurrence in the models.

# Modeling

To begin the modeling process, we split the data into training (75%) and test (25%) data. 
```{r, echo=FALSE}
set.seed(1)
rowTrain <- createDataPartition(y = stroke_data$stroke, p = .75, list = FALSE)

trainData <- stroke_data[rowTrain, ]
testData <- stroke_data[-rowTrain, ]
```

We can examine how many rows contain missing data in the training and test
datasets.
```{r}
nrow_missing_train <- nrow(trainData %>% filter_all(any_vars(is.na(.))))
nrow_missing_test <- nrow(testData %>% filter_all(any_vars(is.na(.))))
```

There are `r nrow_missing_train` rows in the training data containing missing values, and there are `r nrow_missing_test` rows in the test data containing missing values.

## Imputation
We apply the k nearest neighbor (k = 5) method for imputation where we assume any missing data is missing at random.
```{r}
rec <- recipe(stroke ~., data = trainData) %>%
  step_impute_knn(all_predictors(), neighbors = 5) %>%
  step_center(all_numeric_predictors()) %>%
  step_scale(all_numeric_predictors())

prep <- prep(rec, training = trainData)
trainData_pp <- as.data.frame(bake(prep, new_data = trainData))
testData_pp <- as.data.frame(bake(prep, new_data = testData))
```

The training data now contains `r nrow(trainData_pp)` rows. The test data 
now contains `r nrow(testData_pp)` rows.

## Model Fitting
```{r}
ctrl <- trainControl(method = "repeatedcv", 
                     repeats = 5,
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE)
```
The resampling method we used for all of the models is as follows: k-fold cross validation, two class summary (since the ROC curve is only for two classes), and classProbs = T, since AUC/ROC is the evaluation criteria. 

## Penalized Logistic Regression
The first model we fit to the dataset was Penalized Logistic Regression. Some assumptions of logistic regression include requiring the observations to be independent of each other, little or no multicollinearity among independent variables, and linearity of independent variables and log odds. These assumptions are also limitations of the model.
```{r, echo=FALSE}
glmnGrid <- expand.grid(.alpha = seq(0, 1, length = 21), 
                        .lambda = exp(seq(-8, 2, length = 100)))

set.seed(1)
model.glmn <- train(stroke ~ ., 
                    data = trainData_pp, 
                    method = "glmnet",
                    tuneGrid = glmnGrid,
                    metric = "ROC",
                    trControl = ctrl)
```

The tuning parameter in the penalty term of the model controls its flexibility. We decided to test a grid of tuning parameters. For alpha, since we wanted to use the elastic net method, we created a sequence from 0 to 1. For lambda, we created a sequence of numbers to test different values. The model then selects optimal values from the grid which are: alpha = `r model.glmn$bestTune$alpha` and lambda = `r model.glmn$bestTune$lambda`. To display this visually, we created a plot of the model AUC for various tuning parameters in the grid. This plot can be viewed in the appendix.

4 predictors are included in the final model, as seen when the coefficients are extracted.
```{r, echo=FALSE}
coef(model.glmn$finalModel, model.glmn$bestTune$lambda)
```

These include age, having hypertension, having heart disease and average glucose level. This suggests they play important roles in predicting the response.

## MARS
Next, we trained a MARS model on the imputed training data. The MARS model has two tuning parameters: degree and nprune. Degree is the degree of interaction and nprune is the number of retained terms. We decided to test a grid of tuning parameters. 
```{r warning=FALSE, message=FALSE}
set.seed(1)
mars.fit <- train(x = data.frame(trainData_pp[,1:10]),
                  y = pull(trainData_pp, stroke),
                  preProcess = c("center", "scale"),
                  method = "earth",
                  tuneGrid = expand.grid(degree = 1:3,
                                         nprune = 2:22),
                  metric = "ROC",
                  trControl = ctrl)
```

The cross-validation plot can be viewed in the appendix.


The value of degree that maximizes the cross-validated AUC is `r mars.fit$bestTune$degree`, and the value of nprune that maximizes the cross-validated AUC is `r mars.fit$bestTune$nprune`.

The model that maximizes the cross-validated AUC can be viewed in the appendix.

A limitation of MARS models is that they can suffer from high variance. 

## SVM with Linear Kernel
The next model we fit to the imputed training data was a Support Vector Machine with a Linear Kernel. An assumption of the model is that the linearly separable. This is also a limitation of the model, as they don't perform well when this isn’t the case.
```{r include =FALSE}
set.seed(1)
svml.fit <- train(stroke ~ . , 
                  data = trainData_pp,
                  method = "svmLinear",
                  tuneGrid = data.frame(
                    C = exp(seq(-7,2,len = 10))),
                  metric = "ROC",
                  trControl = ctrl)
```

The model contains a tuning parameter C, also known as Cost, that determines the possible misclassifications. It essentially imposes a penalty to the model for making an error: the higher the value of C, the less likely it is that the SVM algorithm will misclassify a point. We decided to test a grid of tuning parameters for C. The model then selects the optimal value from the grid which maximizes the model AUC. That value is C = `r svml.fit$bestTune$C`. To display this visually, we created a plot of the model AUC for various values of the tuning parameter in the grid. This plot can be viewed in the appendix.

The final SVM with a linear kernel that maximizes the cross-validated AUC needed is displayed below.
```{r}
svml.fit$finalModel
```

It needed 421 support vectors to fit the line.

As a reference to understand the model better, let's take a look at a variable importance plot
```{r}
svm_imp <- varImp(svml.fit, scale = F)
plot(svm_imp)
```

## SVM with Radial Kernel

We then trained a Support Vector Machine with a Radial Kernel on the imputed training data. The model has two tuning parameters: C and sigma. C quantifies the cost of misclassification and sigma is related to the flexibility of the decision boundary. We decided to test a grid of tuning parameters.

```{r include = FALSE}
svmr.grid <- expand.grid(C = exp(seq(-2,3,len = 10)),
                         sigma = exp(seq(-8,-1,len = 20)))
set.seed(1)
svmr.fit <- train(stroke ~ . , 
                  data = trainData_pp,
                  method = "svmRadialSigma",
                  tuneGrid = svmr.grid,
                  metric = "ROC",
                  trControl = ctrl)
```

The cross-validation plot can be viewed in the appendix.

The values of C and sigma that maximize the cross-validated AUC are
`r round(svmr.fit$bestTune$C, 3)` and
`r round(svmr.fit$bestTune$sigma, 5)`.

Let's get the support vector machine with radial kernel that maximizes the cross-validated AUC.
```{r}
svmr.fit$finalModel
```

It needed 387 support vectors to fit the model.

A limitation of SVMs is the black box nature of these functions. The use of kernels to separate the data makes them difficult to interpret. SVMs also do not directly provide probability estimates.

## Boosting

```{r}
gbmA.grid <- expand.grid(n.trees = c(2000,3000,4000,5000),
                        interaction.depth = 5:11,
                        shrinkage = c(0.0005,0.001,0.002),
                        n.minobsinnode = 1)

set.seed(1)

#dat <- trainData_pp[1:11]
#install.packages("xgboost")
#library(xgboost)

gbmA.fit <- train(stroke ~ . ,
                  trainData_pp,
                  tuneGrid = gbmA.grid,
                  trControl = ctrl,
                  method = "gbm",
                  distribution = "adaboost",
                  metric = "ROC",
                  verbose = FALSE)

ggplot(gbmA.fit, highlight = TRUE)
```

```{r}
gbmA.pred <- predict(gbmA.fit, newdata = trainData_pp, type = "prob")[,2]


```

```{r}
roc.gbmA <- roc(trainData_pp, gbmA.pred)
plot(roc.gbmA, legacy.axes = TRUE)
```

We will now fit a variable importance plot. 

```{r}
summary(gbmA.fit$finalModel, las = 2, cBars = 19, cex.names = 0.6)
```

With boosting for classification, we are combining weak learners to make a strong learner. In essence, we are repeatedly fitting classification trees to weighted versions of the training data and updating the weights in order to better classify previously misclassified observations. Through this process we are minimizing exponential loss over the whole distribution. 

## Random Forest
```{r}
rf.grid <- expand.grid(mtry = 1:10,
                       splitrule = "gini",
                       min.node.size = 1:6)

set.seed(1)
rf.fit <- train(stroke ~ . ,
                trainData_pp,
                method = "ranger",
                tuneGrid = rf.grid,
                metric = "ROC",
                trControl = ctrl)

```
The cross-validation plot can be viewed in the appendix.

## Comparing Models

To decide which model we will use to predict the response variable, let's use the AUC to compare model performance.
```{r}
resamp <- resamples(list(glmn = model.glmn,
                         mars = mars.fit,
                         svml = svml.fit,
                         svmr = svmr.fit,
                         rf = rf.fit))

tibble(model = c("Penalized Logistic Regression", "MARS", 
                 "SVM (Linear Kernel)", "SVM (Radial Kernel)",
                 "Random Forest"),
       AUC = c(round(mean(unlist(select(resamp$values, "glmn~ROC"))), 3),
               round(mean(unlist(select(resamp$values, "mars~ROC"))), 3),
               round(mean(unlist(select(resamp$values, "svml~ROC"))), 3),
               round(mean(unlist(select(resamp$values, "svmr~ROC"))), 3),
               round(mean(unlist(select(resamp$values, "rf~ROC"))), 3))) %>%
  arrange(desc(AUC)) %>%
  knitr::kable()
```
The penalized logistic regression model has the highest cross-validated AUC
and will be used as the final model.

Let's get its AUC on the imputed test dataset and plot the ROC curve.
```{r}
pred.glmn <- predict(model.glmn, newdata=trainData_pp, type= "prob")[,1]
roc.glmn <- roc(trainData_pp$stroke,pred.glmn)
auc <- c(roc.glmn$auc[1])

modelName <- c("Penalized Logistic Regression")

ggroc(list(roc.glmn),legacy.axes = TRUE) + scale_color_discrete(labels=paste0(modelNames,"(",round(auc,3),")"), name="Model (AUC)") +
  geom_abline(intercept=0,slope=1,color="grey") +
  theme_bw() +
  labs(title = "AUC curve on Imputed Test Data") +
  theme(plot.title = element_text(hjust = 0.5))

```

## Conclusion


## Appendix

### Penalized Logistic Regression Cross-Validation Plot
```{r}
myCol <- rainbow(25)
myPar <- list(superpose.symbol = list(col = myCol),
superpose.line = list(col = myCol))
plot(model.glmn, par.settings = myPar, xTrans = log)
```

### MARS Cross-Validation Plot
```{r}
ggplot(mars.fit) +
  theme_bw() +
  labs(title = "Cross-validated AUC as a function of \n Number of Terms Colored by Degree",
       x = "Number of Terms",
       y = "Cross-validated AUC") +
  theme(plot.title = element_text(hjust = 0.5))
```

### MARS Final Model
```{r}
final_model_mars <- coef(mars.fit$finalModel)
terms_mars <- final_model_mars %>%
  data.frame() %>%
  labels() %>%
  unlist()
tibble(terms = terms_mars[1:5], est = pull(as_tibble(final_model_mars), value)) %>%
  knitr::kable(col.names = c("term", "estimate"))
```

### SVM (Linear Kernel) Cross-Validation Plot
```{r}
plot(svml.fit, highlight = TRUE, xTrans = log)
```

### SVM (Radial Kernel) Cross-Validation Plot
```{r}
myCol <- rainbow(20)
myPar <- list(superpose.symbol = list(col = myCol),
              superpose.line = list(col = myCol))
plot(svmr.fit, highlight = TRUE, par.settings = myPar)
```

### Random Forest Cross-Validation Plot
```{r}
ggplot(rf.fit, highlight = TRUE)
```

